prolfqua - R-package for Analysing Proteomics Label Free Quantification Experiments

### 2022-03-17 inst/MyArticle/paper.Rmd
Merging benchmark figures into 1.

---
title: prolfqua - R-package for Analysing Proteomics Label Free Quantification Experiments
author: 
  - name: Witold Wolski
    affil: 1; 2
    orcid: '0000-0002-6468-120X'
  - name: Paolo Nanni
    affil: 1
    orcid: '0000-0001-8429-3557'
  - name: Jonas Grossmann
    affil: 1; 2
    orcid: '0000-0002-6899-9020'
  - name: Maria D'Errico
    affil: 1; 2
    orcid: '0000-0001-9467-9058'
  - name: Ralph Schlapbach
    affil: 1
    orcid: '0000-0002-7488-4262'
  - name: Christian Panse
    affil: 1; 2
    orcid: '0000-0003-1975-3064'
affiliation:
  - num: 1
    address: Functional Genomics Center Zurich (https://www.fgcz.ch/)
  - num: 2
    address: Swiss Institute of Bioinformatics (https://www.sib.swiss/)
column_numbers: 5
poster_height: "30in"
poster_width: "58in"
title_textsize:	"90pt"
font_family: Monaco
primary_colour: "#000C66"
titletext_fontfamily: Arial
body_textsize: "35px"
logoleft_name: C:/Users/wolski/__checkout/prolfqua/inst/Figures/FGCZLogo.png
logoright_name: C:/Users/wolski/__checkout/prolfqua/inst/Figures/hexStickerProlfqua.png
output: 
  bookdown::html_document2:
    toc: true
  posterdown::posterdown_html:
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
---

@@

We were inspired by the R package caret [@caret2008], which enables to call various machine learning methods, making selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the analysis of LFQ data. However, after examining the R packages for modeling proteomics quantification data, we observed that model specification, input, and output formats differ widely and wildly. However, all of them fit some type of linear model either to peptide or protein intensities, determine contrasts among conditions, and afterward, some of them apply empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular design (TODO - clarity), with R linear models as a core, where features such as p-value aggregation (ROPECA) or variance shrinkage could be easily enabled if required.

Models in _prolfqua_ are specified using R’s linear model and mixed model formula interface. Therefore, knowledge of the R regression model infrastructure (TODO cite) is an advantage when using our package. Furthermore, this glass box approach should make it easy to reimplement an analysis performed with _prolfqua_ using base R or other programming languages by reading the analysis script and without looking at the package code. However, acknowledging the R formula interface’s complexity to non-statisticians and the popularity of _MSstats_, we provide the functionality to suggest a model formula from the sample annotation.

An important aspect of mass spectrometric data are missing values. Missing values lead to groups of varying size. Linear and mixed effect models can handle unbalanced designs, i.e. different group sizes. As long as there is at least one observation per group they do produce unbiased estimates. We use this feature instead of imputing missing observations. However, if there are not observation in one of the groups, then contrasts involving this group can not be estimated. To address this, for the group without observations, we use a point prior equal to the average of $1\%$ smallest group mean intensities in the dataset and infer the variance from the variance in other groups. (TODO use different start of sentence) A way to address this is to annotate the produced models with information about missingness which is done by for instance _MSStats_. How missingness can modelled using the Hurdle model is described here (TODO citations), however to our knowledge only the R package _proDA_ models missingness directly.

@@

```{r LFQData, echo=FALSE, fig.cap="Class Diagram for LFQData and related objects. The LFQData_Plotter class uses the LFQData class to implement methods for plotting.", out.width = '90%'}
fig_svg <- cowplot::ggdraw() + cowplot::draw_image("LFQDataUML.svg")
plot(fig_svg)

@@

```{r ContrastUML, echo=FALSE, fig.cap="(ref:ContrastUML)", out.width = '90%'}
fig_svg <- cowplot::ggdraw() + cowplot::draw_image("ContrastClassesUML.svg")
plot(fig_svg)
```

@@

```{r prepro,  echo = TRUE , fig.cap="(ref:scaling)", out.width = '90%', fig.width = 8, fig.height = 8}
library(prolfqua)
d <- prolfqua_data('data_ionstar')$filtered()
lfqd <- LFQData$new(d$data, d$config) 
t <- lfqd$get_Transformer()
lfqd <-  t$log2()$robscale()$lfq
pl <- lfqd$get_Plotter()
p_a <- pl$intensity_distribution_density() +
  ggplot2::labs(tag = "A") + ggplot2::theme(legend.position = "none")
agr <- lfqd$get_Aggregator()
lfqp <- agr$medpolish()
p_b <- agr$plot()$plots[[1]] + ggplot2::labs(tag = "B")
p_c <- lfqp$get_Stats()$violin() + ggplot2::labs(tag = "C")
pl <- lfqp$get_Plotter()
p_d <- pl$boxplots()$boxplot[[1]] + ggplot2::labs(tag = "D")
ggpubr::ggarrange(p_a, p_b, p_c, p_d)

@@

```{r exampleContrasts, fig.cap="(ref:exampleContrasts)", echo=TRUE, out.width = '90%'}
contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c"
  #,"dilution_(6/4.5)_1.3(3)" =   "dilution.c - dilution.b",
  #,"dilution_(4.5/3)_1.5" =   "dilution.b - dilution.a"
)
# fit model
lmmodel <- paste(lfqp$intensity_column()," ~ dilution.")
modelFunction <- strategy_lm( lmmodel, model_name = "lm")
models <- build_model(lfqp, modelFunction)
p0 <- models$anova_histogram()
# compute contrasts from linear model and with imputation
contr <- prolfqua::Contrasts$new(models, contrasts) |> 
  prolfqua::ContrastsModerated$new()
conI <- prolfqua::ContrastsSimpleImpute$new( lfqp, contrasts) |> 
  prolfqua::ContrastsModerated$new()
tmp <- prolfqua::addContrastResults(prefer = contr, add = conI)
# visualize results
pl <- tmp$merged$get_Plotter()
p1 <- pl$histogram()$p.value
p2 <- pl$volcano()$FDR +
   ggplot2::theme(legend.position = "bottom")
gridExtra::grid.arrange(p1, p2, ncol = 2)

@@

The Benchmark functionality of _prolfqua_ includes ROC curves and computes partial areas under those curves (pAUC) and other scores, e.g. the false discovery proportion FDP. We use it to study how well linear, mixed effect models or p-value moderation model quantitative mass spectrometric high throughput experiments and compare it with results produced by MSstats [@MSstats2014] and by proDA [@bioxrvproDA2020]. Figure \@ref(fig:benchmarkROC)

@@

(TODO(CP): benchmark load from package - first build package with vignettes)
```{r}

allBenchmarks <- readRDS("../../inst/Benchresults/allBenchmarks.RDS")
benchmark_msstats <- readRDS("../../inst/Benchresults/benchmark_msstats.RDS")
benchmark_prodA <- readRDS("../../inst/Benchresults/benchmark_medpolish_proDA.RDS")
msFragger <- readRDS(file = "../../inst/Benchresults/MSFragger_medpol_benchmark.RDS")

@@

(ref:benchmarkROC) Left panel - Partial area under the ROC curve at $10\%$ FPR for all contrasts. Red line average area under the curve. Left panel - Difference to mean partial area under the ROC curve for various models, at $10\%$ FPR, by fold change.

@@

p1 <- ggplot2::ggplot(resAllB, ggplot2::aes(x = Name, y = pAUC_10)) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::facet_wrap(~whatfix)  + 
  ggplot2::coord_cartesian(ylim = c(min(resAllB$pAUC_10),max(resAllB$pAUC_10))) + 
  ggplot2::theme_minimal() + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, vjust = 0.5)) +
  ggplot2::geom_hline(ggplot2::aes(yintercept = meanpAUC_10), color = "red") + ggplot2::xlab("")

p2 <- ggplot2::ggplot(res, ggplot2::aes(x = contrast, y = pAUC_10n, group = Name)) +
  ggplot2::geom_line(stat = "identity", ggplot2::aes(linetype = Name, color = Name), size = 1) + 
  ggplot2::facet_wrap(~ whatfix, scales = "free") +
  ggplot2::theme_minimal() + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, vjust = 0.5)) +
  ggplot2::geom_hline(ggplot2::aes(yintercept = 0), color = "red") + 
   ggplot2::theme(legend.position = "bottom", legend.title = ggplot2::element_blank())

gridExtra::grid.arrange(p1, p2, ncol = 2)

@@

Figure \@ref(fig:benchmarkROC) left panel, shows how various estimates obtained from the models, i.e. $\log_2(FC)$, $t-statistics$ and scaled p-values allow to identify true positives (TP) given a false positive rate. When testing for differential expression, the $\log_2(FC)$ is outperformed by than the $t$-statistic or $p$-value. 
If an accurate estimate of the $log_2(FC)$ is essential, the linear models fitted to protein intensities, estimated using Tukey's median polish, show the best performance. We speculate that this is because of the robustness of Tukey's median polish method. This hypothesis could be examined, by including other forms of protein intensity inferences implemented in `prolfqua`, e.g. `top-N` or `rlm`, into the benchmark.

There are only minor differences in the p-AUC when using the t-statistics or the scaled p-value  (see Figure \@ref(fig:benchmarkROC) right panel). In order to compute the confusion matrices based on the p-value we first need to rescale it (see Equation \@ref(eq:scalepvalue)). Thus, the p-value is a function of the t-statistics and the degrees of freedom. Interestingly, the pAUC increases when using p-values instead of the t-statistics for linear models, while it decreases for mixed effect models. The reason is erroneous denominator degrees of freedom estimation for many proteins in the mixed-effect models.

@@

(ref:FDRfdp) Left panel - Number and percentage of estimated contrasts by modeling method. Right panel - Compare FDR estimate with false discovery proportion (FDP).

@@

dd <- purrr::map_df(allBenchmarks, function(x){res <- x$smc$summary; res$name <- x$model_name;res})
dd
dd <- dd |> dplyr::mutate(nrcontrasts = protein_Id * (4 - nr_missing))
dd
dds <- dd |> dplyr::group_by(name) |> dplyr::summarize(nrcontrasts = sum(nrcontrasts))
dds
dds$percent <- dds$nrcontrasts/max(dds$nrcontrasts) * 100

@@

gridExtra::grid.arrange(p3, p4, ncol = 2)

@@

[Goes to discussion] The method used to infer protein intensities from more peptide intensities is an additional important factor. The original _proDA_ publication uses MaxLFQ [@cox2014accurate] protein estimates. When using those, _proDA_ performs significantly worse than _proDA_ when using the protein estimates obtained from peptide intensity estimates using Tukey's median polish,  $pAUC_10(t-statistics) = 66%$ and $pAUC_10(t-statistics) = 73%$ respectively (see Articles on [@prolfquaGithubIO] for more details).

@@

The probabilistic dropout analysis implemented in the _proDA_ produces inferences comparable to other methods (Figure \@ref(fig:benchmarkROC) for the largest number of proteins (Figure \@ref(fig:FDRfdp) left panel). Moreover, the estimated fold changes are unbiased and show high diagnostic accuracy. Furthermore, the scaled p-values and t-statics are on par with those obtained by the linear model with variance moderation. Therefore, we plan to integrate the _proDA_ package as an additional modeling option into _prolfqua_ (see Figure \@ref(fig:ContrastUML)).

>>>>

---
title: prolfqua - R-package for Analysing Proteomics Label Free Quantification Experiments
author: 
  - name: Witold Wolski
    affil: 1; 2
    orcid: '0000-0002-6468-120X'
  - name: Paolo Nanni
    affil: 1
    orcid: '0000-0001-8429-3557'
  - name: Jonas Grossmann
    affil: 1; 2
    orcid: '0000-0002-6899-9020'
  - name: Maria D'Errico
    affil: 1; 2
    orcid: '0000-0001-9467-9058'
  - name: Ralph Schlapbach
    affil: 1
    orcid: '0000-0002-7488-4262'
  - name: Christian Panse
    affil: 1; 2
    orcid: '0000-0003-1975-3064'
affiliation:
  - num: 1
    address: Functional Genomics Center Zurich (https://www.fgcz.ch/)
  - num: 2
    address: Swiss Institute of Bioinformatics (https://www.sib.swiss/)
output: 
  bookdown::pdf_document2:
    toc: true
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
---

@@

We were inspired by the R package caret [@caret2008], which enables to call various machine learning methods, making selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the analysis of LFQ data. However, after examining the R packages for modeling proteomics quantification data, we observed that model specification, input, and output formats differ widely and wildly. However, all of them fit some type of linear model either to peptide or protein intensities, determine contrasts among conditions, and afterward, some of them apply empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object oriented design, with R linear models as a core, where features such as p-value aggregation (ROPECA) or variance shrinkage could be easily enabled (see Figure \@ref(fig:ContrastUML)).

Models in _prolfqua_ are specified using R’s linear model and mixed model formula interface. Therefore, knowledge of the R regression model infrastructure [@faraway2016extending,@MASS2002] is an advantage when using our package. Furthermore, this glass box approach should make it easy to reimplement an analysis which was performed with _prolfqua_ using base R or other programming languages by reading the analysis script. However, acknowledging the R formula interface’s complexity to non-statisticians and the popularity of _MSstats_, we provide the functionality to suggest a model formula from the sample annotation provided in tabular form similarily to MSStats [@Choi2014].

An important aspect of mass spectrometric data are missing values. Missing values lead to groups of varying size. Linear and mixed effect models can handle unbalanced designs, i.e. different group sizes. As long as there is at least one observation per group they do produce unbiased estimates. We use this feature instead of imputing missing observations. 
However, if there are no observation in one of the groups, then contrasts involving this group can not be estimated. To address this, for the group without observations, we use a point prior equal to the average of $3\%$ smallest group intensities means in the dataset and infer the variance using the pooled the variance of other groups. Proteins for which the fold change was imputed in such a way are explicitely annotated in the result table.

Other means of modelling missing information in proteomics data using the Hurdle model is discussed by [@goeminne2020msqrob], while the R package _proDA_ models missingness using probabilistic dropout models [@bioxrvproDA2020].

@@

(ref:LFQData) Class Diagram for LFQData and related classes. The LFQData_Plotter class uses the LFQData class to implement methods for plotting. Similarily the LFQData_Stats, LFQData_Summary, reference the LFQData class and group methods for variance and sample size estimation or for summarizing peptide and protein counts.

```{r LFQData, echo=FALSE, fig.cap="(ref:LFQData)", out.width = '90%'}

knitr::include_graphics("LFQDataUML.png")

@@

```{r ContrastUML, echo=FALSE, fig.cap="(ref:ContrastUML)", out.width = '90%'}
#fig_svg <- cowplot::ggdraw() + cowplot::draw_image("ContrastClassesUML.svn")
#plot(fig_svg)

knitr::include_graphics("ContrastClassesUML.png")
```

@@

```{r prepro,  echo = TRUE , fig.cap="(ref:scaling)", out.width = '90%', fig.width = 8, fig.height = 8}
library(prolfqua)
d <- prolfqua_data('data_ionstar')$filtered()
lfqd <- LFQData$new(d$data, d$config) 
t <- lfqd$get_Transformer()
lfqd <-  t$log2()$robscale()$lfq
pl <- lfqd$get_Plotter()
p_a <- pl$intensity_distribution_density() +
  ggplot2::labs(tag = "A") + ggplot2::theme(legend.position = "none")
agr <- lfqd$get_Aggregator()
lfqp <- agr$medpolish()

p_b <- agr$plot()$plots[[1]] + ggplot2::labs(tag = "B")
p_c <- lfqp$get_Stats()$violin() + ggplot2::labs(tag = "C")
pl <- lfqp$get_Plotter()
p_d <- pl$boxplots()$boxplot[[1]] + ggplot2::labs(tag = "D")
ggpubr::ggarrange(p_a, p_b, p_c, p_d)

@@

```{r exampleContrasts, fig.cap="(ref:exampleContrasts)", echo=TRUE, out.width = '90%'}
contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c"
)
# fit model
lmmodel <- paste(lfqp$intensity_column()," ~ dilution.")
modelFunction <- strategy_lm( lmmodel, model_name = "lm")
models <- build_model(lfqp, modelFunction)
p0 <- models$anova_histogram()
# compute contrasts from linear model and with imputation
contr <- prolfqua::Contrasts$new(models, contrasts) |> 
  prolfqua::ContrastsModerated$new()
conI <- prolfqua::ContrastsSimpleImpute$new( lfqp, contrasts) |> 
  prolfqua::ContrastsModerated$new()
tmp <- prolfqua::addContrastResults(prefer = contr, add = conI)
# visualize results
pl <- tmp$merged$get_Plotter()
p1 <- pl$histogram()$p.value
p2 <- pl$volcano()$FDR +
   ggplot2::theme(legend.position = "bottom")
gridExtra::grid.arrange(p1, p2, ncol = 2)

@@

The Benchmark functionality of _prolfqua_ includes ROC curves and computes partial areas under those curves (pAUC) and other scores, e.g. the false discovery proportion FDP. We use it to study how well linear, mixed effect models or p-value moderation model quantitative mass spectrometric high throughput experiments and compare it with results produced by MSstats [@MSstats2014] and by proDA [@bioxrvproDA2020]. Figure \@ref(fig:FDRfdp)

@@

```{r}
allBenchmarks <- readRDS("../../inst/Benchresults/allBenchmarks.RDS")
benchmark_msstats <- readRDS("../../inst/Benchresults/benchmark_msstats.RDS")
benchmark_prodA <- readRDS("../../inst/Benchresults/benchmark_medpolish_proDA.RDS")
msFragger <- readRDS(file = "../../inst/Benchresults/MSFragger_medpol_benchmark.RDS")

@@

(ref:benchmarkROC)  Left panel - Difference to mean partial area under the ROC curve for various models, at $10\%$ FPR, by fold change.

@@

p1_pAUC <- ggplot2::ggplot(resAllB, ggplot2::aes(x = Name, y = pAUC_10)) +
  ggplot2::geom_bar(stat = "identity") +
  ggplot2::facet_wrap(~whatfix)  + 
  ggplot2::coord_cartesian(ylim = c(min(resAllB$pAUC_10),max(resAllB$pAUC_10))) + 
  ggplot2::theme_minimal() + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, vjust = 0.5)) +
  ggplot2::geom_hline(ggplot2::aes(yintercept = meanpAUC_10), color = "red") + ggplot2::xlab("")

p2_compare_variousLevels <- ggplot2::ggplot(res, ggplot2::aes(x = contrast, y = pAUC_10n, group = Name)) +
  ggplot2::geom_line(stat = "identity", ggplot2::aes(linetype = Name, color = Name), size = 1) + 
  ggplot2::facet_wrap(~ whatfix, scales = "free") +
  ggplot2::theme_minimal() + 
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, vjust = 0.5)) +
  ggplot2::geom_hline(ggplot2::aes(yintercept = 0), color = "red") + 
  ggplot2::theme(legend.position = "bottom", legend.title = ggplot2::element_blank())

@@

Figure \@ref(fig:FDRfdp) left panel, shows how various estimates obtained from the models, i.e. $\log_2(FC)$, $t-statistics$ and scaled p-values allow to identify true positives (TP) given a false positive rate. When testing for differential expression, the $\log_2(FC)$ is outperformed by than the $t$-statistic or $p$-value. 
If an accurate estimate of the $log_2(FC)$ is essential, the linear models fitted to protein intensities, estimated using Tukey's median polish, show the best performance. We speculate that this is because of the robustness of Tukey's median polish method. This hypothesis could be examined, by including other forms of protein intensity inferences implemented in `prolfqua`, e.g. `top-N` or `rlm`, into the benchmark.

There are only minor differences in the p-AUC when using the t-statistics or the scaled p-value  (see Figure \@ref(fig:FDRfdp) right panel). In order to compute the confusion matrices based on the p-value we first need to rescale it (see Equation \@ref(eq:scalepvalue)). Thus, the p-value is a function of the t-statistics and the degrees of freedom. Interestingly, the pAUC increases when using p-values instead of the t-statistics for linear models, while it decreases for mixed effect models. The reason is erroneous denominator degrees of freedom estimation for many proteins in the mixed-effect models.

@@

(ref:FDRfdp) Left panel - Partial area under the ROC curve at $10\%$ FPR for all contrasts and three different statistics: the difference among conditions, the scaled p-value (sign(diff) * p.value) and the t-statics (higher is better). The red line average area under the curve. Left panel - Number and percentage of estimated contrasts for each modeling method (higher is better). Right panel - Compare FDR estimate with false discovery proportion (FDP) closer to the diagnoal (red continous line is better).

@@

dd <- purrr::map_df(allBenchmarks, function(x){res <- x$smc$summary; res$name <- x$model_name;res})
dd <- dd |> dplyr::mutate(nrcontrasts = protein_Id * (4 - nr_missing))
dds <- dd |> dplyr::group_by(name) |> dplyr::summarize(nrcontrasts = sum(nrcontrasts))
dds$percent <- dds$nrcontrasts/max(dds$nrcontrasts) * 100

@@

hlay = rbind(c(1,1),
            c(2,3))

gridExtra::grid.arrange(p1_pAUC, p3, p4, layout_matrix=hlay)

@@

[Goes to discussion] The method used to infer protein intensities from more peptide intensities is an additional important factor. The original _proDA_ publication uses MaxLFQ [@cox2014accurate] protein estimates. When using those, _proDA_ performs significantly worse than _proDA_ when using the protein estimates obtained from peptide intensity estimates using Tukey's median polish,  $pAUC_{10}(\textrm{t-statistics}) = 66%$ and $pAUC_{10}(\textrm{t-statistics}) = 73%$ respectively (see Articles on [@prolfquaGithubIO] for more details).

@@

The probabilistic dropout analysis implemented in the _proDA_ produces inferences comparable to other methods (Figure \@ref(fig:FDRfdp) for the largest number of proteins (Figure \@ref(fig:FDRfdp) left panel). Moreover, the estimated fold changes are unbiased and show high diagnostic accuracy. Furthermore, the scaled p-values and t-statics are on par with those obtained by the linear model with variance moderation. Therefore, we plan to integrate the _proDA_ package as an additional modeling option into _prolfqua_ (see Figure \@ref(fig:ContrastUML)).
